---
title: "Figure 4c<br>UMAP - NMF Dimensionality Reduction"
editor_options: 
  chunk_output_type: console
---

### Required R Packages
```{r}
#| label: packages
#| warning: false
#| message: false
.libPaths("/packages/")

# Load packages
pacman::p_load(
  matrixStats,
  tidyverse,
  data.table,
  fs,
  qs,
  Matrix,
  RcppML,
  ragg
)
```

### Project Setup and Custom Functions
```{r}
#| label: project setup and custom functions
#| warning: false

# Define project directories
project_dir <- path("~/HCMI-Flagship-Manuscript")
r_data_dir <- path(project_dir, "data/DNA_methylation")
out_dir <- path(project_dir, "plots")

# Source custom functions
source(path(r_data_dir,
            "HCMI_DNA_methylation_analysis_custom_functions_colors.R"))

# Set random seed for reproducibility
set.seed(1022)
```

### Data Loading: DNA Methylation Beta Value Matrix
- DNA methylation beta value matrix including HCMI, TCGA, and TARGET samples
- Available on the HCMI publication page (25.78 GB in size)
```{r}
#| label: beta value matrix
fname <- "HCMI_TCGA_TARGET_DNA_methylation_beta_value_matrix"
dt <- qread(path(r_data_dir, fname, ext = "qs"),
            nthreads = parallelly::availableCores() - 1)

dim(dt)
```

### Data Loading: Sample Annotations and Metadata
- Sample annotations and metadata for HCMI, TCGA, and TARGET samples
```{r}
#| label: sample annotations
fname <- "HCMI_TCGA_TARGET_DNA_methylation_analysis_sample_table"
dt_sample <- readRDS(path(r_data_dir, fname, ext = "rds"))
dim(dt_sample)
# Validate sample IDs match between matrices
if (!identical(colnames(dt)[-1], dt_sample[, sample_id])) {
  stop("Sample IDs do not match between beta matrix and sample table")
}
```

### Sample Filtering
- Filter samples based on:
  - CpG probe success rate (>90%)
  - Model purity (Based on DNA analysis)
  - Expanded models (Duplicates of original models)
```{r}
#| label: sample filtering
dt_sample <- dt_sample[cpg_probe_success_rate > 0.9 &
                         dna_analysis_low_purity_exclusion == FALSE &
                         !(sample_type %like% "Expanded")]
dim(dt_sample)
sample_ids <- dt_sample[, sample_id]
dt <- dt[, .SD, .SDcols = c("ProbeID", sample_ids)]
dim(dt)
```

### Probe Filtering
- Filter out problematic probes and probes on sex chromosomes, and mitochondrial DNA
- Probe manifest available at https://zhouserver.research.chop.edu/InfiniumAnnotation
```{r}
#| label: probe filtering problematic probes
fname <- "EPICv1_manifest_hg38"
dt_probe <- readRDS(path(r_data_dir, fname, ext = "rds"))
dt_probe <- dt_probe[MASK_general == FALSE &
                       probeType == "cg" &
                       !(seqnames %in% c("chrX", "chrY", "chrM"))]
probe_ids <- dt_probe[, ProbeID]
dt <- dt[ProbeID %in% probe_ids]
dim(dt)
```

### Cancer-Associated Probes
- Using hypermethylated probes identified in `scripts/analyses/Epigenetic_Concordance_Fig.3a.qmd`
```{r}
#| label: cancer-associated probes
fname <- "HCMI_tumor_model_hypermethylated_probes"
probe_ids <- readRDS(path(r_data_dir, fname, ext = "rds"))
length(probe_ids)
dt <- dt[ProbeID %in% probe_ids]
dim(dt)
```

### Additional Probe Filtering
- Filter based on:
  - Missing values (<10%)
  - Hypermethylation frequency
```{r}
#| label: probe filtering missing values
dt_stat <- probe_stats(data = dt, me_cutoff = 0.3)
dt_stat <- dt_stat[perc_na < 10 & num_meth > 1]
probe_ids <- dt_stat[, probe_id]
dt <- dt[ProbeID %in% probe_ids]
dim(dt)
gc(verbose = FALSE)
``` 

### NMF Preprocessing
- Prepare matrix for Non-negative Matrix Factorization
- Handle zeros and missing values
```{r}
#| label: NMF preprocessing
df <- as.matrix(dt, rownames = "ProbeID")
# Handle zeros and missing values
df[df == 0] <- 1e-12
df[is.na(df)] <- 0
```

### NMF Rank Selection via Cross-Validation
- Evaluate optimal ranks for NMF using cross-validation
```{r}
#| label: NMF cv
#| eval: false
cv <- RcppML::crossValidate(df,
                            k = c(seq(20, 150, by = 10),
                                  seq(155, 200, by = 5)),
                            method = "predict", reps = 3,
                            verbose = FALSE)

# Save the results
fname <- "HCMI_DNA_methylation_NMF_CV_results"
saveRDS(cv, file = path(r_data_dir, fname, ext = "rds"))
```

### Cross-Validation Results Visualization
- Plot for evaluating optimal rank
```{r}
#| label: cv plot
fname <- "HCMI_DNA_methylation_NMF_CV_results"
cv <- readRDS(path(r_data_dir, fname, ext = "rds"))
p <-  ggplot(cv, aes(x = k, y = value, color = rep)) +
  geom_point() +
  geom_line() +
  scale_x_continuous(limits = c(0, 200),
                     breaks = seq(0, 200, by = 10)) +
  theme_fig4c +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.95),
        axis.text.y  = element_text(colour = "black", size = 12))
p
```

### NMF Model Training
- Train rank-170 NMF model
```{r}
#| label: NMF rank-170 model
#| eval: false
model <- RcppML::nmf(df,
                     k = 170,
                     seed = 1:10,
                     tol = 1e-6,
                     verbose = TRUE)
```                     

### Save the Trained NMF Model
```{r}
#| label: save
#| eval: false
fname <- "HCMI_DNA_methylation_NMF_k170_model"
saveRDS(model, file = path(r_data_dir, fname, ext = "rds"))
```

### Session Information
```{r}
#| label: R session
sessionInfo()
```