---
title: "Figure 4e<br>Joint Heatmap Analysis of HCMI and TCGA samples"
editor_options: 
  chunk_output_type: console
---

### Required R Packages
```{r}
#| label: packages
#| warning: false
#| message: false
.libPaths("/packages/")

# Load packages
pacman::p_load(
  matrixStats,
  tidyverse,
  data.table,
  fs,
  qs,
  doFuture,
  ComplexHeatmap,
  viridis,
  circlize,
  ragg
)
```

### Project Setup and Custom Functions
```{r}
#| label: project setup and custom functions
#| warning: false

# Define project directories
project_dir <- path("~/HCMI-Flagship-Manuscript")
r_data_dir <- path(project_dir, "data/DNA_methylation")
out_dir <- path(project_dir, "plots")

# Source custom functions and colors
source(path(r_data_dir,
            "HCMI_DNA_methylation_analysis_custom_functions_colors.R"))
```

### Data Loading: DNA Methylation Beta Value Matrix
- DNA methylation beta value matrix including HCMI, TCGA, and TARGET samples
- Available on the HCMI publication page (25.78 GB in size)
```{r}
#| label: beta value matrix
fname <- "HCMI_TCGA_TARGET_DNA_methylation_beta_value_matrix"
dt <- qread(path(r_data_dir, fname, ext = "qs"),
            nthreads = parallelly::availableCores() - 1)
dim(dt)
```

### Data Loading: Sample Annotations and Metadata
- Sample annotations and metadata for HCMI, TCGA, and TARGET samples
```{r}
#| label: sample annotations
fname <- "HCMI_TCGA_TARGET_DNA_methylation_analysis_sample_table"
dt_sample <- readRDS(path(r_data_dir, fname, ext = "rds"))
dim(dt_sample)
# Verify sample IDs match between beta matrix and sample table
if (!identical(colnames(dt)[-1], dt_sample[, sample_id])) {
  stop("Sample IDs do not match between beta matrix and sample table")
}
```

### Sample Filtering
- Filter samples based on:
  - CpG probe success rate (>90%)
  - Model purity (Based on DNA analysis)
  - Expanded models (Duplicates of original models)
```{r}
#| label: sample filtering
dt_sample <- dt_sample[cpg_probe_success_rate > 0.9 &
                         dna_analysis_low_purity_exclusion == FALSE &
                         !(sample_type %like% "Expanded")]
sample_ids <- dt_sample[, sample_id]
dt <- dt[, .SD, .SDcols = c("ProbeID", sample_ids)]
dim(dt)
```

### Sample Selection by Cancer Type
- Select samples from four major cancer types
```{r}
#| label: four major cohorts
gbm_ids <- dt_sample[tcga_hcmi_code == "GBM" &
                       tmp_hcmi_code == "LGGGBM", sample_id]
stad_ids <- dt_sample[tmp_hcmi_code %in% "STADESCA", sample_id]
paad_ids <- dt_sample[tcga_hcmi_code == "PAAD" &
                        tmp_hcmi_code == "PAAD", sample_id]
coad_ids <- dt_sample[tmp_hcmi_code %in% "COADREAD", sample_id]
sample_ids <- c(gbm_ids, stad_ids, paad_ids, coad_ids)
dt_sample <- dt_sample[sample_id %in% sample_ids]
```

### Sample Selection by TMP Subtype Availability
```{r}
#| label: TMP subtypes availability
dt_sample <- dt_sample[!is.na(tmp_subtype)]
sample_ids <- dt_sample[, sample_id]
dt <- dt[, .SD, .SDcols = c("ProbeID", sample_ids)]
dim(dt)
```

### Probe Filtering
- Filter out problematic probes and probes on sex chromosomes, and mitochondrial DNA
- Probe manifest available at https://zhouserver.research.chop.edu/InfiniumAnnotation
```{r}
#| label: probe filtering
fname <- "EPICv1_manifest_hg38"
dt_probe <- readRDS(path(r_data_dir, fname, ext = "rds"))
dt_probe <- dt_probe[MASK_general == FALSE &
                       probeType == "cg" &
                       !(seqnames %in% c("chrX", "chrY", "chrM"))]
probe_ids <- dt_probe[, ProbeID]
dt <- dt[ProbeID %in% probe_ids]
dim(dt)
gc(verbose = FALSE)
```

### Cancer-Associated Probes
- Using hypermethylated probes identified in `scripts/analyses/Epigenetic_Concordance_Fig.3a.qmd`
```{r}
#| label: cancer-associated probes
fname <- "HCMI_tumor_model_hypermethylated_probes"
probe_ids <- readRDS(path(r_data_dir, fname, ext = "rds"))
dt <- dt[ProbeID %in% probe_ids]
dim(dt)
```

### Probe Selection Based on DNA Methylation Variability
```{r}
#| label: probe selection by cancer type
tmp_hcmi_codes <- c("LGGGBM", "COADREAD", "STADESCA", "PAAD")

# Check sample counts by cancer type
dt_sample[, .N, tmp_hcmi_code]

# Select probes based on DNA methylation variability for each cancer type
var_probe_ids <- foreach(i = seq_along(tmp_hcmi_codes)) %do% {
  sample_ids <- dt_sample[tmp_hcmi_code == tmp_hcmi_codes[i], sample_id]
  dt1 <- dt[, .SD, .SDcols = c("ProbeID", sample_ids)]
  dt_stat <- probe_stats(data = dt1, me_cutoff = 0.3)
  dt_stat <- dt_stat[perc_na < 10 & num_meth > 1]
  sel <- nrow(dt_stat) * 0.1
  dt_stat <- dt_stat[order(-probe_sds)][seq_len(sel)]
  probe_ids <- dt_stat[, probe_id]
}
probe_ids <- purrr::reduce(var_probe_ids, union)
length(probe_ids)
dt <- dt[ProbeID %in% probe_ids]
gc(verbose = FALSE)
```

### Binarize the beta values and perform hierarchical clustering on the samples
```{r}
#| label: binarizing and clustering
df <- as.matrix(dt, rownames = "ProbeID")
df_bin <- df >= 0.3
mode(df_bin) <- "numeric"
d <- dist(t(df_bin), method = "binary", diag = TRUE)
fit <- hclust(d, method = "ward.D2")
dt_fit <- data.table(sample_id = colnames(df)[fit$order],
                     hcc_order = 1:ncol(df))
dt_sample <- dt_sample[dt_fit, on = "sample_id"]
```

### Define Sample Order
```{r}
#| label: sample order
dt_sample <- dt_sample[order(tmp_hcmi_code, tmp_subtype, hcc_order)]
sample_ids <- dt_sample[, sample_id]
df <- df[, sample_ids]
#dim(df)
cancer_type_levels <- c("LGGGBM", "STADESCA", "COADREAD", "PAAD")
cancer_types <- factor(dt_sample[, tmp_hcmi_code], levels = cancer_type_levels)
```

### Heatmap
- Sample annotations bars
```{r}
#| label: heatmap column annotations

# Verify sample IDs match between beta matrix and sample table
if (!identical(dt_sample[, sample_id], colnames(df))) {
  stop("Sample IDs do not match between beta matrix and sample table")
}

column_ha <- HeatmapAnnotation(
  Project = dt_sample[, project],
  Sample_type = dt_sample[, sample_type_simple],
  Subtype = dt_sample[, tmp_subtype],
  col = list(
    Project = project_colors,
    Sample_type = sample_type_colors,
    Subtype = tmp_subtype_colors
  ),
  annotation_name_gp = gpar(fontsize = 6.5),
  show_legend = FALSE
)
```
- Heatmap
```{r}
#| label: heatmap plot preparation
ht <- Heatmap(df,
  cluster_columns = FALSE,
  column_split = cancer_types,
  cluster_column_slices = FALSE,
  column_gap = unit(0.5, "mm"),
  show_column_dend = FALSE,
  show_column_names = FALSE,
  cluster_rows = TRUE,
  show_row_dend = FALSE,
  show_row_names = FALSE,
  col = col_fun_turbo100,
  top_annotation = column_ha,
  show_heatmap_legend = TRUE,
  column_title = str_c(ncol(df), " samples"),
  row_title = str_c(nrow(df), " CpGs"),
  use_raster = FALSE
)
```
```{r}
#| label: heatmap plot
#| fig-align: left
#| dpi: 300
#| fig-format: png
#| out-width: "80%"
#| out-height: "80%"
ht
```
- Save in a png file
```{r}
#| label: heatmap save
#| eval: false
fname <- "Fig.4e"
agg_png(file = path(out_dir, fname, ext = "png"),
        background = "transparent",
        scaling = 0.75,
        res = 300,
        width = 7.5,
        height = 5,
        units = "in")
ht <- draw(ht, background = "transparent")
dev.off()
```

### Session Information
```{r}
#| label: R session
sessionInfo()
```