---
title: "Extended Data Figure 5c<br>UMAP - Visualization of the NMF Model (All Samples)"
editor_options: 
  chunk_output_type: console
---

### Required R Packages
```{r}
#| label: packages
#| warning: false
#| message: false
.libPaths("/packages/")

# Load packages
pacman::p_load(
  tidyverse,
  data.table,
  fs,
  RcppML,
  umap,
  ggnewscale,
  ragg
)
```

### Project Setup and Custom Functions
```{r}
#| label: project setup and custom functions
#| warning: false

# Define project directories
project_dir <- path("~/HCMI-Flagship-Manuscript")
r_data_dir <- path(project_dir, "data/DNA_methylation")
out_dir <- path(project_dir, "plots")

# Source custom functions and colors
source(path(r_data_dir,
            "HCMI_DNA_methylation_analysis_custom_functions_colors.R"))

# Set random seed for reproducibility
set.seed(1022)
```

### Data Loading: NMF Model
```{r}
#| label: NMF H matrix

# Load NMF model
fname <- "HCMI_DNA_methylation_NMF_k170_model"
nmf_model <- readRDS(path(r_data_dir, fname, ext = "rds"))

# Extract H matrix and convert to data.table format
nmf_h <- as.data.table(nmf_model$h, keep.rownames = TRUE)
setnames(nmf_h, "rn", "ProbeID")


```






### Data Loading: Sample Annotations and Metadata
- Sample annotations and metadata for HCMI, TCGA, and TARGET samples
```{r}
#| label: sample annotations
fname <- "HCMI_TCGA_TARGET_DNA_methylation_analysis_sample_table"
dt_sample <- readRDS(path(r_data_dir, fname, ext = "rds"))
dim(dt_sample)
#dt_sample
```

### Sample Filtering
- Filter samples based on:
  - CpG probe success rate (>90%)
  - Model purity (Based on DNA analysis)
  - Expanded models (Duplicates of original models)
```{r}
#| label: sample filtering
dt_sample <- dt_sample[cpg_probe_success_rate > 0.9 &
                         dna_analysis_low_purity_exclusion == FALSE &
                         !(sample_type %like% "Expanded")]
dim(dt_sample)

# Verify sample IDs match between NMF H matrix and sample table
if (!identical(colnames(nmf_h)[-1], dt_sample[, sample_id])) {
  stop("Sample IDs do not match between NMF H matrix and sample table")
}

# Check sample counts by project
dt_sample[, .N, project]
```

### UMAP
```{r}
#| label: UMAP

# Prepare data for UMAP
nmf_ht <- data.table::transpose(nmf_h,
                                make.names = "ProbeID",
                                keep.names = "sample_id")
nmf_ht <- as.matrix(nmf_ht, rownames = "sample_id")

# Run UMAP
nmf_umap <- umap(nmf_ht,
                 metric = "cosine",
                 n_neighbors = 15,    # Default value
                 min_dist = 0.1,      # Default value
                 verbose = FALSE)

# Convert results to data.table format
nmf_umap <- data.table(nmf_umap$layout, keep.rownames = TRUE)
setnames(nmf_umap, c("sample_id", "umap1", "umap2"))

# Merge with sample metadata
nmf_umap <- dt_sample[nmf_umap, on = "sample_id"]

# Reorder factor levels
nmf_umap[, project_tumor_model := fct_relevel(project_tumor_model,
                                              c("TCGA_TUMOR",
                                                "TARGET_TUMOR",
                                                "HCMI_TUMOR",
                                                "HCMI_MODEL"))]
nmf_umap <- nmf_umap[order(project_tumor_model)]
```
- Remove TCGA outliers [KIRP samples with CpG island methylator phenotype (CIMP)]
```{r}
#| label: TCGA outliers
nmf_umap[umap2 > 30][, .N, .(project, tmp_hcmi_code, tmp_subtype_name)]
nmf_umap <- nmf_umap[umap2 < 30]
```
- Save UMAP coordinates (as provided in Supplementary Table 11)
```{r}
#| label: umap coordinates
#| eval: false
nmf_umap_coord <- nmf_umap[, .SD,
                           .SDcols = c("sample_id",
                                       "umap1",
                                       "umap2",
                                       "tcga_hcmi_code",
                                       "project_tumor_model")]
fname <- "Extended_Data_Fig.5c_DNA_methylation_UMAP_coordinates"
write.table(nmf_umap_coord,
            file = path(out_dir, fname, ext = "csv"),
            sep = ",",
            row.names = FALSE,
            quote = FALSE)
```

### Visualization
```{r}
#| label: plot preparation

# Base plot
p <- ggplot(data = nmf_umap, aes(x = umap1, y = umap2)) +
  # TCGA/TARGET samples
  geom_point(data = nmf_umap[project_tumor_model %in%
                               c("TCGA_TUMOR", "TARGET_TUMOR")],
             size = 1.8, shape = 0, alpha = 0.5,
             aes(color = tcga_hcmi_code, stroke = 0.1)) +
  labs(x = "UMAP 1", y = "UMAP 2") +
  scale_color_manual(values = hcmi_tcga_colors) +
  new_scale_color() +
  # HCMI samples
  geom_point(data = nmf_umap[project_tumor_model %like% "HCMI"],
             size = 1.8, stroke = 0.075,
             aes(shape = project_tumor_model,
                 color = project_tumor_model,
                 fill = tcga_hcmi_code,
                 alpha = project_tumor_model)) +
  scale_shape_manual(values = shape_map) +
  scale_alpha_manual(values = alpha_map) +
  scale_color_manual(values = color_map_border) +
  scale_fill_manual(values = hcmi_tcga_colors) +
  theme_fig4c +
  theme(
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_text(size = 14, face = "bold"),
        axis.line = element_line(colour = "black", linewidth = 0.17),
        legend.position = "none")
```
```{r}
#| label: plot
#| fig-align: left
#| dpi: 300
#| fig-format: png
#| out-width: "70%"
#| out-height: "70%"
p
```
- Save in a png file
```{r}
#| label: save
#| eval: false
fname <- "Extended_Data_Fig.5c"
ggsave(file = path(out_dir, fname, ext = "png"),
       p,
       device = agg_png,
       scaling = 1,
       dpi = 300,
       width = 5.5,
       height = 5.5,
       units = "in")
```

### Session Information
```{r}
#| label: R session
sessionInfo()
```
