---
title: Epigenetic Concordance, Related to Figure 3a
editor_options: 
  chunk_output_type: console
---

### Required R Packages
```{r}
#| label: packages
#| warning: false
#| message: false
.libPaths("/packages/")

# Load packages
pacman::p_load(
  matrixStats,
  tidyverse,
  data.table,
  fs,
  doFuture
)
```

### Project Setup and Custom Functions
```{r}
#| label: project setup and custom functions
#| warning: false

# Define project directories
project_dir <- path("~/HCMI-Flagship-Manuscript")
r_data_dir <- path(project_dir, "data/DNA_methylation")
out_dir <- path(project_dir, "plots")

# Source custom functions and colors
source(path(r_data_dir,
            "HCMI_DNA_methylation_analysis_custom_functions_colors.R"))

# Configure parallel processing
future::plan(multisession, workers = 4)
options(future.globals.maxSize = 10000 * 1024^2)

# Set random seed for reproducibility
set.seed(1022)
```

### Data Loading: DNA Methylation Beta Value Matrix
- DNA methylation beta value matrix available on the HCMI publication page in CSV format (4.38 GB in size)
```{r}
#| label: beta value matrix
fname <- "HCMI_DNA_methylation_beta_value_matrix"
dt <- readRDS(path(r_data_dir, fname, ext = "rds"))
dim(dt)
```

### Data Loading: Sample Annotations and Metadata
```{r}
#| label: sample annotations
fname <- "HCMI_DNA_methylation_analysis_sample_table"
dt_sample <- readRDS(path(r_data_dir, fname, ext = "rds"))
dim(dt_sample)

# Validate sample IDs match between matrices
if (!identical(colnames(dt)[-1], dt_sample[, sample_id])) {
  stop("Sample IDs do not match between data matrix and sample table")
}
```

### Sample Filtering
- Filter samples based on:
  - CpG probe success rate (>70%)
  - Model purity (Based on DNA analysis)
- Create a concatenated string of matched tumor and model aliquot IDs for later use
```{r}
#| label: sample filtering
dt_sample <- dt_sample[cpg_probe_success_rate > 0.7 &
                         dna_analysis_low_purity_exclusion == FALSE]
sample_ids <- dt_sample[, sample_id]
dt <- dt[, .SD, .SDcols = c("ProbeID", sample_ids)]

# Matched Tumor-Model Aliquot IDs
dt_sample[, tumor_model_aliquotes := str_c(matched_tumor_aliquot,
                                           cancer_model_aliquot,
                                           sep = "_")]
dt_sample[, tumor_exp_model_aliquotes := str_c(matched_tumor_aliquot,
                                               matched_expanded_model_aliquot,
                                               sep = "_")]
paired_aliquots_ids <- dt_sample[
  !is.na(tumor_model_aliquotes),
  tumor_model_aliquotes
]

paired_aliquots_ids2 <- dt_sample[
  !is.na(tumor_exp_model_aliquotes),
  tumor_exp_model_aliquotes
]
paired_aliquots_ids <- unique(c(paired_aliquots_ids, paired_aliquots_ids2))
length(paired_aliquots_ids)
#head(paired_aliquots_ids)
```

### Probe Filtering
- Filter out problematic probes and probes on sex chromosomes, and mitochondrial DNA
- Probe manifest available at https://zhouserver.research.chop.edu/InfiniumAnnotation
```{r}
#| label: probe filtering
fname <- "EPICv1_manifest_hg38"
dt_probe <- readRDS(path(r_data_dir, fname, ext = "rds"))
dim(dt_probe)
dt_probe <- dt_probe[MASK_general == FALSE &
                       probeType == "cg" &
                       !(seqnames %in% c("chrX", "chrY", "chrM"))]
probe_ids <- dt_probe[, ProbeID]
dt <- dt[ProbeID %in% probe_ids]
dim(dt)
```

### Probes Unmethylated in Normal Tissues
- Provided in Supplementary Table 11
- See methods for details on how these probes were identified
```{r}
#| label: probes unmethylated in normal tissues
fname <- "HCMI_normal_tissue_unmethylated_probes"
probe_ids <- readRDS(path(r_data_dir, fname, ext = "rds"))
length(probe_ids)
dt <- dt[ProbeID %in% probe_ids]
dim(dt)
gc(verbose = FALSE)
```

### Identifying Cancer-Associated Probes
- Minimum threshold of 0.3 used to identify hypermethylated CpGs for each cancer type
```{r}
#| label: cancer-associated probes
diagnosis_code <- dt_sample[, .N, diagnosis_tcga_hcmi_code][order(-N)]
setnames(diagnosis_code, c("code", "num"))
codes <- diagnosis_code[, code]

probe_ids_ls <- foreach(i = seq_along(codes)) %do% {
  code <- codes[i]
  num <- diagnosis_code[i, num]
  sample_ids <- dt_sample[diagnosis_tcga_hcmi_code %in% code, sample_id]
  dt_diagnosis  <- dt[, .SD, .SDcols = c("ProbeID", sample_ids)]
  dt_stat <- probe_stats(data = dt_diagnosis, me_cutoff = 0.3)

  # Apply sample size-dependent filtering
  if (num >= 2) {
    dt_stat <- dt_stat[num_meth >= 2]
  } else {
    dt_stat <- dt_stat[num_meth > 0]
  }
  dt_stat[order(num_meth)]
  probe_ids <- dt_stat[, probe_id]
}
names(probe_ids_ls) <- codes
diseases_cohorts <- names(probe_ids_ls)
probe_ids <- purrr::reduce(probe_ids_ls, union)
length(probe_ids)
```
- Save for later use
```{r}
#| label: save
fname <- "HCMI_tumor_model_hypermethylated_probes"
saveRDS(probe_ids, file = path(r_data_dir, fname, ext = "rds"))
```

### Additional Diseases/Samples Filtering
- Exclude cancer types that lack either a matched tumor or model sample
```{r}
#| label: additional sample filtering
# Tumor types with only one sample
diagnosis_codes1 <- dt_sample[, .N, diagnosis_tcga_hcmi_code][
  N == 1, diagnosis_tcga_hcmi_code
]
diagnosis_codes1
diseases_cohorts <- diseases_cohorts[!diseases_cohorts %in% diagnosis_codes1]
# Tumor only or model only
diagnosis_codes1 <- c("IPMN", "NHL", "SSCC", "SIAD", "THCA")
diseases_cohorts <- diseases_cohorts[!diseases_cohorts %in% diagnosis_codes1]
```

### Model-Tumor Concordance Calculation
- Function to calculate Pearson correlations between model and tumor samples for a specific disease type
```{r}
#| label: function
#' Calculate Pearson correlations between tumor and model samples for a specific disease
#' @param disease Character string indicating the disease code
#' @param dt Data table containing methylation beta values
#' @param dt_sample Data table containing sample annotations
#' @param probe_ids_ls List of probe IDs to use for each disease
#' @param paired_aliquots_ids Vector of paired aliquot IDs
#' @return Data table with correlation results
calculate_disease_correlations <- function(
  disease,
  dt,
  dt_sample,
  probe_ids_ls,
  paired_aliquots_ids
) {
  # Get disease-associated probes
  probe_ids <- probe_ids_ls[[disease]]
  num_probes <- length(probe_ids)

  # Get sample IDs
  model_ids <- dt_sample[diagnosis_tcga_hcmi_code == disease &
                           tumor_model_type %like% "Model", sample_id]
  tumor_ids <- dt_sample[tumor_model_type == "Tumor", sample_id]

  # Subset data for this disease
  dt_disease <- dt[ProbeID %in% probe_ids]

  # Create all model-tumor pairs
  pairs <- CJ(model = model_ids, tumor = tumor_ids)

  # Calculate correlations in parallel
  dt_cor <- foreach(i = seq_len(nrow(pairs)), .combine = rbind) %dofuture% {
    model <- pairs$model[i]
    tumor <- pairs$tumor[i]

    dt_pair <- dt_disease[, .SD, .SDcols = c(model, tumor)]
    cor_r <- as.numeric(cor(dt_pair[, ..model], dt_pair[, ..tumor],
                            method = "pearson",
                            use = "pairwise.complete.obs"))

    data.table(
      Model_aliquot = model,
      Tumor_aliquot = tumor,
      Pearson_correlation = cor_r,
      Diagnosis_code_model = disease,
      Num_of_cpgs_examined = num_probes,
      Tumor_model_aliquotes = str_c(tumor, model, sep = "_")
    )
  }

  # Add disease codes for tumors
  tumor_aliquots <- dt_cor[, Tumor_aliquot]
  diagnosis_code_tumors <- dt_sample[
    match(tumor_aliquots, sample_id),
    diagnosis_tcga_hcmi_code
  ]
  dt_cor[, Diagnosis_code_tumor := diagnosis_code_tumors]

  # Add case IDs and group labels
  dt_cor[, `:=`(
    Case_id_model = str_extract(
      Model_aliquot,
      "HCM-[:alnum:]{4}-[:alnum:]{4}-[:alnum:]{3}"
    ),
    Case_id_tumor = str_extract(
      Tumor_aliquot,
      "HCM-[:alnum:]{4}-[:alnum:]{4}-[:alnum:]{3}"
    ),
    Group = "different_cohort"
  )]

  dt_cor[
    Diagnosis_code_model == Diagnosis_code_tumor,
    Group := "same_cohort"
  ]

  dt_cor[
    Tumor_model_aliquotes %in% paired_aliquots_ids,
    Group := "matched_case_same_cohort"
  ]

  return(dt_cor)
}
```
- Run the function over all diseases
```{r}
#| label: calculate correlations for all diseases
# Initialize progress tracking
pb <- txtProgressBar(min = 0, max = length(diseases_cohorts), style = 3)

# Calculate correlations for each disease
dt_cor_all <- foreach(disease = diseases_cohorts, .combine = rbind) %do% {
  setTxtProgressBar(pb, which(disease == diseases_cohorts))

  result <- calculate_disease_correlations(
    disease,
    dt,
    dt_sample,
    probe_ids_ls,
    paired_aliquots_ids
  )

  result
}

close(pb)
```
- Clean up and organize the results
```{r}
#| label: clean up and organize final results
dt <- dt_cor_all[, .(
  Model_aliquot,
  Tumor_aliquot,
  Pearson_correlation,
  Num_of_cpgs_examined,
  Group,
  Diagnosis_code_model,
  Diagnosis_code_tumor
)]
dt[, Group := factor(Group,
                     levels = c("matched_case_same_cohort",
                                "same_cohort",
                                "different_cohort"))]
dt <- dt[order(Diagnosis_code_model,
               Group,
               Model_aliquot)]
dt[, .N, Group]
dt
```
- Save the results
```{r}
#| label: output
#| eval: false
fname <- "DNA_methylation_concordance_fig.3a"
write.table(dt,
            file = path(out_dir, fname, ext = "csv"),
            sep = ",",
            row.names = FALSE,
            col.names = TRUE,
            quote = FALSE)
```

### Session Information
```{r}
#| label: R session
sessionInfo()
```
